### 2020-02-24 thesis/paper.tex
grammar

For example, Castellanos-Garz\'{o}n and D\'{ı}az1 \cite{DNAVis} propose ``An Evolutionary and Visual Framework for Clustering of DNA Microarray Data'' which implements specific visualizations of the data that may not be useful when using other data than DNA data. Still with this knowledge and this specific task, these visualizations may prove more useful than only the ones targeted at the general use-case of clustering. There are various examples for domains in which such specific frameworks exist, even for the task of clustering ``Memes'' \cite{Dang2017} to detect emerging events and topics driven by rumours, a framework has been created.

@@

\section{Consensus Clustering}\label{sec:consensus}
Consensus clustering aims to combine multiple clustering solutions for the same data-set into one result. This result should be representative of the combined knowledge acquired by all base clusterings and therefor be more robust and accurate than the average result at least. As the consensus result should lie in the center of the solution space spanned by the base clusterings, if there are sufficiently many good clusterings it should be able to overcome problems arising from outliers. In literature the problem of consensus clustering is often found under different names, those being cluster ensemble \cite{BOONGOEN20181}, clustering aggregation \cite{Gionis2005ClusteringA} and ensemble of partitions \cite{ensemblepartitions}. For the evaluation of the quality of a result five main factors can be defined \cite{Ghaemi2009ASC,survey1}:\newline

@@

\subsection{Object Co-occurrence}
Object Co-occurrence (O-Co) based consensus functions look at how points relate to each other and use this information to create a consensus result. If two points often occur in a common cluster of the base clusterings intuition would suggest that they should also appear in a common cluster in the final result. The same holds for points most frequently clustered apart, indicating that they should not be clustered together by the consensus function. The different O-Co based methods formulate algorithms that aim to create a result reflecting these thoughts and usually run in a deterministic way, not seeing this task as an optimization problem. This contrasts the methodology of Median Partitioning methods as can be seen in the following section. To summarize the groups of O-Co consensus functions they are categorized by their underlying computational strategy. The different methods are based on:

@@

Relabeling and Voting based methods use the individual points as  ``voters'' whereby each point can vote for which cluster it should be assigned to \cite{4470298}. For this to work, the different base clusterings need to map their clusters to one another, defining which cluster is associated most closely with which cluster from another base clustering. This mapping can be performed with algorithms like Kuhn’s Hungarian method \cite{Kuhn2010}.

Co-association Matrix based method create a matrix representing how often two points are in the same cluster relatively  \cite{Monti2003}. With this matrix the consensus result can be computed using hierarchical linking of sets or with a clustering algorithm that uses this matrix as a similarity matrix. The Link Uncertainty \cite{6413733} and Dual-Similarity \cite{7344797} methods expand on this idea by allowing uncertainty values in the matrix and using neighbor-neighbor relationships respectively. Spectral Methods also expand on the idea of co-association matrices by trying to remove noise from the matrix before creating the final result \cite{Tao:2016:RSE:2983323.2983745}.

Graph and Hypergraph based methods create graphs from the base clusterings, representing the relationship between the points as edges. They then split the resulting graph into sub-graphs with graph partitioning \cite{Strehl:2003:CEK:944919.944935}, defining the clusters. Representative methods from this group are the Cluster-based Similarity Partitioning Algorithm (CSPA), the HyperGraph-Partitioning Algorithm (HGPA) and the Meta-CLustering Algorithm (MCLA). A variant of these are Minimum Spanning Tree algorithms which require a specific graph structure, a minimum spanning tee. This tree is created using with Prim's Algorithm \cite[p.~276]{10.1007/978-3-540-85033-5_27} and the clusters are obtained by cutting edges. With $k$ edges cut the final result is derived through majority voting, a prominent method of this category is Divisive Clustering Ensemble with Automatic Cluster Number (DICLENS) \cite{6035671}. DICLENS is also able to estimate an optimal $k$ by computing all possible results and evaluating their quality.

Locally Adaptive Clustering Algorithm look at different subspaces of the data and compute weights from clustering results in these subspaces \cite{Domeniconi2007}. To create a final result, the computed clusterings and their weights are modeled as a graph, which is then partitioned.

@@

\subsection{Median Partitioning}
The main difference of Median Partitioning in comparison to Object Co-occurrence is that methods in this group tackle consensus clustering as an optimization problem. With the distance between clusterings defined the task of finding a median partition is to look for a solution that would be in the center of the solution space spanned by the base clusterings. In other words the resulting clustering should have a minimal sum of distances to all base clusterings. As this problem has been shown to be NP-hard \cite{np_median_partition}, heuristics must be used. The different groups of median partitioning methods can be roughly described through the methods that they are based on:

@@

Genetic Algorithm based methods require a distance measure and a fitness function to compute a consensus result \cite{Cristofor2002FindingMP}. Through evolutionary steps different results, are combined or modified until the fitness of the result is satisfactory or does not further improve over many iterations. At first the results to be combined may be picked from the base clusterings or just be random assignments. For the choice of distance measure, Mirkin Distance is often used and comes with additional ideas on how genetic modifications should be applied \cite{5766165}, such that a good result is computed with higher probability. The Sum of Pairwise Distances can also be used with genetic algorithms. The assumption of this measure is that a good consensus result can be found by more strongly considering clusterings far apart in the solution space \cite{6694095}. The idea of using Kernel Functions was also proposed \cite{Vega-Pons:2010:WPC:1786814.1787121}, enabling faster computation regardless of the heuristic used, though mostly genetic algorithms are used with this aswell.

Non-negative Matrix Factorization based methods translate the median partitioning problem to a matrix form which can be solved for by running two multiplicative update procedures \cite{Li:2007:SCS:1441428.1442121}. Running those procedures multiple times iteratively improves the result each iteration and the final result can be obtained by stopping the algorithm at any time or whenever some convergence criterion is met.

Binary Linear Programming based methods create a compact representation of the data and model objects on an intermediate level between points and clusters \cite{HUANG2016131}. For those intermediate objects binary decisions on cluster ownership are calculated iteratively using Expectation Maximization (EM), whereby within the expectation step of EM the factor graph method \cite{HUANG2016131} is used for estimation. The solution is then mapped back from intermediate objects to data points, assigning the points their labels for the consensus solution.

\part{Theory}
\chapter{Methods}\label{cha:methods}
The newly created tool uses a large amount of existing methods which are described in this chapter. Here, I aim to give an overview of what exact methods were used, the theory behind them and in which scenarios they can be/are applied. For a more general overview of some of the topics relating the methods used, refer to the related work of this paper in Chapter \ref{cha:related_work}.

\section{Dimensionality Reduction}\label{sec:dim_reduction}
Dimensionality reduction techniques are able to reduce the number of dimensions in a data-set while trying to maintain as much information as possible. Reducing the number of dimensions is often needed for reducing the computational complexity of clustering and improving the accuracy of the result, but also to visualize the data in a lower-dimensional space. This visualization is then often easier to understand for a user than if all dimensions are shown at once.
\subsection{Principal Component Analysis}
 Principal Component Analysis (PCA) \cite{pca} aims to transform correlated variables into uncorrelated ones thereby omitting less relevant dimensions. Generally, Principal Components (PCs) are linear combinations of the original variables and can be computed through the eigenvectors $V$ of the covariance matrix $C$. To use these PCs for dimensionality reduction, the $k$ onces corresponding to the $k$ largest eigenvalues of the matrix are extracted into a matrix $V_k$ and used as a basis to transform the data. This is done by calculating $data_{reduced}=V_k^T*data_{center}$, whereby the original data must be centered first.

\subsection{t-Distributed Stochastic Neighbor Embedding}
The dimensionality recution method t-Distributed Stochastic Neighbor Embedding (t-SNE) \cite{Maaten2008VisualizingDU} works by converting the Euclidean distances between points into conditional probabilities that represent similarities. This similarity represents the probability that a point would be picked as the neighbor of another point if neighbors were picked proportionally to their probability density under a Student-t distribution centered at a given point. To lay out the points, an initial randomly sampled solution is created and improved on iteratively with gradient decent, such that the similarities are best represented by the points in low-dimensional space. As inputs for the method, the number of dimensions $k$, the perplexity and some optimization parameters are needed. The number of resulting dimensions is defined by $k$ and the optimization parameters can be used to improve the runtime of the algorithm. The perplexity parameter can be seen as a smoothing parameter defining the effective number of neighbors for each point. This parameter is usually also fairly robust \cite{Maaten2008VisualizingDU}, good choices in practice are values between $5$ and $50$.

\subsection{Multi-Dimensional Scaling}
Multi-Dimensional Scaling (MDS) \cite{mds} uses a distance matrix (often called proximity data) and a number of target dimensions $k$ as input. It then trys to lay out the points in a $k$-dimensional space such that the distance between each pair in that space is as close as possible to its true distance. When this method is used with a distance matrix created with Euclidean distances, the result will be equivalent to simply running the PCA method on the data itself.

For the calculation of the solution, MDS builds $n$ $k$-dimensional hypervectors with unknown parameters and aims to minimize the stress value by applying gradient decent. The stress of a configuration is defined as the residual sum of squares for the distances, i.e. the sum of squared errors between the actual distances and the distances in the $k$-dimensional space.

@@

\subsection{k-Means}
K-Means clustering is one of the most prominent methods used for clustering \cite{Jin2010}. The main implementations of k-Means are Lloyds method \cite{4031353} and MacQueens method \cite{macqueen1967} although important variants like k-Medians \cite{Dohan2015KmedianAT} and k-Medoids \cite{kmedoids}  are available.

@@

The main difference between Lloyds and MacQueens method is that Lloyds method only recalculates the centroid positions every time an iteration is completed while MacQueens method also updates them whenever a point changes its assigned clusters. This results in more computational cost during iterations, may allow for faster convergence though. The k-Medians method is different in regards to the distance metric used. Unlike Lloyds and MacQueens methods it does not use the Euclidean distance but the Manhattan-distance (1-Norm) to find the centroids. K-Medoids in contrast to all previous methods does not use centroids that may lie anywhere in the data space, but only existing points may be used. These are in this case called medoids and define the objects for which the average dissimilarity to all points within the cluster is minimal.

\subsection{Expectation Maximization}
The Expectation Maximization (EM) algorithm \cite{10.2307/2984875} aims to iteratively find estimates of parameters of a model such that they are assigned the most likely values. To do this, the algorithm alternates between two steps, those being the expectation step and the maximization step. In the expectation step the expectation of the log-likelihood is defined as a function given the current set of parameters. The maximization step then computes the new set of parameters such that the expected log-likelihood function is maximized.

@@

\subsection{DBSCAN}
Density-Based Spatial Clustering of Application with Noise (DBSCAN) \cite{10.5555/3001460.3001507}, is an algorithm that unlike k-Means is not based on partitioning the data directly but on searching for dense regions in the data. For the calculation, two parameters are needed, namely epsilon ($\epsilon$) and a minimum number of objects ($\mu$ or MinPts). The parameter $\epsilon$ defines how large a region may be considered for neighborhood search and $\mu$ defines how many objects must be within that region for it to be considered dense. The following explanations are needed for full understanding of the algorithm:
\newline

When defining the neighborhood within the epsilon range for an object $p$ as $N_\epsilon(p)$, the algorithm considers an object $p$ as a core object if $|N_\epsilon(p)|\geq \mu$.

An object $p$ is said to be directly density-reachable if $p\in N_\epsilon(q) $ and $q$ is a core object.

@@

Any objects at the border are not core objects, but will be assigned to a cluster; points outside will be considered noise.
\newline

@@

The output of the OPTICS clustering algorithm is an ordered list of objects with their reachabililty-distances. To compute this list all points are processed iteratively starting with any point. All points neighboring the current point are updated with their reachability distance from the current point, assigning the minimum of the new and previous values to the point. To select the next point the one with the smallest reachability value is chosen or a new random one if there are none. The resulting list of points with reachability distances can be visualized with a bar chart whereby valleys indicate clusters. As the definition of these valleys can be done in different ways, the result can be considered hierarchical.

\section{Consensus Functions}
Consensus functions have the goal of merging multiple different clustering solutions together such that the overall result represents a consensus between all base clusterings. In this work I focused on two methods that aim to achieve this.
\subsection{Co-association Matrix based Consensus}
Co-association Matrix based consensus functions \cite{Monti2003} first create a co-association matrix that indicates how often two points $p$ and $q$ appear in in a common cluster, i.e. if $p$ and $q$ are clustered together $70\%$ of the time the corresponding entry of the matrix is $0.7$. To extract a consensus result from this matrix different methods can be applied. One possibility is to choose a clustering algorithm that takes this matrix as an input, converts it to a dissimilarity/distance measure and performs clustering with it. Another possibility is to use linking of sets strategies, which is also done for the implementation in the tool. This has the benefit of being trivial to implement and not imposing a cluster structure, as would for example be the case when using k-Means. The possible linking strategies are single-link, using the similarity of the two points of the sets where it is maximal, average-link, using the average similarity between all point pairs, or maximum link, whereby the smallest similarity is used. With each of these strategies the pair of sets for which this linking value is maximal (i.e. they are most similar) is merged. To stop the merging process and obtain the final result either a threshold value can be set for the link value upon which no more merging should be performed or a number of target clusters $k$ can be defined, stopping when only $k$ sets remain. Another option for defining when to stop merging is the Lifetime Criterion. For the implementation of the tool the average-link strategy is chosen per default with either defining $k$ or using this Lifetime Criterion such that is estimates $k$ itself.

@@

\subsection{Kuhn’s Hungarian method}
Kuhn’s Hungarian method \cite{Kuhn2010} can be used to solve the mathematical assignment problem for the two-dimensional case. A simple example for the assignment problem is to think of $n$ jobs and $n$ workers in a factory that each demand a different payment for each specific job. Now the task is to find the optimal assignment whereby all jobs are performed, all workers are assigned a job and the total cost for all workers is minimal. The assignment problem also allows for reformulation such that the target value is maximized instead of minimized, whenever this better fits the problem and it can be extended to more than two dimensions. In the aforementioned example an additional dimension could for example be the tools used, i.e. a worker demands a different payment depending on which task he has to perform with which tool being provided. Also, as an extension to make this more general each dimension may not necessarily need to have the same number of elements. If there are for example more workers than jobs, some workers will not be assigned a job.

@@

\subsection{Kernel Density Estimation}
Kernel Density Estimation aims to estimate the probability density function for a random variable \cite{parzen1962estimation}. Given a sample of points one is interested in the shape of the unknown density function of the distribution from which the points are drawn. To estimate this function a kernel density estimator is used, requiring a kernel function and a bandwidth parameter. For the kernel function many different choices are possible, typical ones are Gaussian, multivariate Student and Laplacian kernels \cite{kim2011robust}. The bandwidth parameter defines the smoothness of the resulting function and can be difficult to manually set. For this reason, estimates based on the standard deviation of the samples are available for this parameter \cite{rulethumbdens}.

@@

\part{Implemented Tool}
\chapter{Using the Tool}\label{cha:Tool}
The implemented tool aims to facilitate the computation and selection of good clustering solutions for a given data-set. To do this, both powerful visualizations and algorithms are needed, which are combined here. When working with the created application the user can first import and edit point data, then select clustering algorithms and parameters for meta-clustering and finally use the meta-view to analyze the results or create a consensus results. The final goal after all of this is either having found a clustering solution that is satisfactory or created one using consensus clustering in the final step. This simple workflow can be described with three steps, each with its own view and functionality. In the following sections the views and possibilities within each view are described. For more details on implementation see Chapter \ref{cha:impl}.

\section{Data Visualization and Manipulation}
To start with the tool, the user needs to import or create a data-set on which the clustering should be performed. As the data may have too many dimensions or include unnecessary columns, it needs to be sanitized as well. For this reason, the first window shown implements functionality for both of this steps. When the application is started a scatter plot without data is shown, as well as a side menu with different buttons. This starting window will further on also be called data-view as it visualizes point data, it can be seen in Figure \ref{fig:data-view}.

@@

The scatter plot in the center left shows data that was imported or created and allows for changing the shown dimensions by clicking on the center of the corresponding axis. The range of values may also be adjusted this way, clicking on the upper or lower bound of the range on any axis opens a small input box. To automatically set this range, such that the minimum and maximum values of the shown data are just at the border, the small square button in the bottom left can be clicked. This is also a useful feature for returning to a full view of the data if the range was made smaller to have a zoomed view of the data. The right part of the tool contains the functionality for importing/exporting data, manipulating the data, showing it in a scatter plot matrix and opening the workflow-view which will be explained in the next step.

 Firstly, to import a data-set for visualization and later use the import button in the bottom right can be selected, opening a file dialog which can load either CSV or ARFF files. These formats are supported as they are most commonly used to store such data. This file dialog can be seen in Figure \ref{fig:data-import}.

@@

After selecting a file the index of cluster labels can be defined if labels should be loaded from the file. Otherwise, the value can be left at $-1$ indicating that no columns should be used for label information. ``$-2$'' may be also be specified, it is used as a shortcut for selecting the last column as cluster label information, removing the need to count columns if it is the last one. The chosen column is then not inserted as a dimension in the data space but used only for the cluster labels, which the color of the loaded points will reflect in the scatter plot of the data-view. In Figure \ref{fig:loaded-data}, one can see how the data-view looks when data with cluster labels is shown. The data-set used for this example is the same one as in Section \ref{sec:unob}.

@@

To export data the export button can be used. This button will open a similar dialog as for importing, though the index of the labels does not need to be defined, as data with cluster labels will be stored such that the labels for all points will be included in the first column. If the data does not contain label information, this first column is completely omitted. The supported format for exporting data is CSV.

Alternatively to importing, the user can also create a data-set himself from within the tool. It is possible to draw points onto the canvas when the ``SinglePoint'' option is selected in the top right, as in Figure \ref{fig:data-view}. For this, the user only needs to click on the position of the canvas where he wants points placed. It is important to note that if there are more than two dimensions all values for the not shown dimensions will not be set, which may cause a problem for the clustering algorithms later on. Another possibility is the use of the ``ELKIGenerator'', which accepts a XML-style input for defining how to generate data. The definition for this format can be found in corresponding definitions file on the ELKI Github repository \cite{elkixml} or in the tools lib folder. This generator allows defining distributions on dimensions, the number of points for each generated distribution and many additional options like plane rotations. This can be quite useful when wanting to work with a quick sample data-set.

>>>>

For example, Castellanos-Garz\'{o}n and D\'{ı}az1 \cite{DNAVis} propose ``An Evolutionary and Visual Framework for Clustering of DNA Microarray Data'' which implements specific visualizations of the data that may not be useful when using other data than DNA data. Still, with this knowledge and this specific task, these visualizations may prove more useful than only the ones targeted at the general use-case of clustering. There are various examples for domains in which such specific frameworks exist, even for the task of clustering ``Memes'' \cite{Dang2017} to detect emerging events and topics driven by rumours, a framework has been created.

@@

\section{Consensus Clustering}\label{sec:consensus}
Consensus clustering aims to combine multiple clustering solutions for the same data-set into one result. This result should be representative of the combined knowledge acquired by all base clusterings and therefore be more robust and accurate than the average result at least. As the consensus result should lie in the center of the solution space spanned by the base clusterings if there are sufficiently many good clusterings it should be able to overcome problems arising from outliers. In literature the problem of consensus clustering is often found under different names, those being cluster ensemble \cite{BOONGOEN20181}, clustering aggregation \cite{Gionis2005ClusteringA} and ensemble of partitions \cite{ensemblepartitions}. For the evaluation of the quality of a result, five main factors can be defined \cite{Ghaemi2009ASC,survey1}:\newline

@@

\subsection{Object Co-occurrence}
Object Co-occurrence (O-Co) based consensus functions look at how points relate to each other and use this information to create a consensus result. If two points often occur in a common cluster of the base clusterings, intuition would suggest that they should also appear in a common cluster in the final result. The same holds for points most frequently clustered apart, indicating that they should not be clustered together by the consensus function. The different O-Co based methods formulate algorithms that aim to create a result reflecting these thoughts and usually run in a deterministic way, not seeing this task as an optimization problem. This contrasts the methodology of Median Partitioning methods as can be seen in the following section. To summarize the groups of O-Co consensus functions they are categorized by their underlying computational strategy. The different methods are based on:

@@

Relabeling and Voting based methods use the individual points as  ``voters'' whereby each point can vote for which cluster it should be assigned to \cite{4470298}. For this, to work, the different base clusterings need to map their clusters to one another, defining which cluster is associated most closely with which cluster from another base clustering. This mapping can be performed with algorithms like Kuhn’s Hungarian method \cite{Kuhn2010}.

Co-association Matrix based methods create a matrix representing how often two points are in the same cluster relatively  \cite{Monti2003}. With this matrix the consensus result can be computed using hierarchical linking of sets or with a clustering algorithm that uses this matrix as a similarity matrix. The Link Uncertainty \cite{6413733} and Dual-Similarity \cite{7344797} methods expand on this idea by allowing uncertainty values in the matrix and using neighbor-neighbor relationships respectively. Spectral Methods also expand on the idea of co-association matrices by trying to remove noise from the matrix before creating the final result \cite{Tao:2016:RSE:2983323.2983745}.

Graph and Hypergraph based methods create graphs from the base clusterings, representing the relationship between the points as edges. They then split the resulting graph into sub-graphs with graph partitioning \cite{Strehl:2003:CEK:944919.944935}, defining the clusters. Representative methods from this group are the Cluster-based Similarity Partitioning Algorithm (CSPA), the HyperGraph-Partitioning Algorithm (HGPA) and the Meta-CLustering Algorithm (MCLA). A variant of these are Minimum Spanning Tree algorithms which require a specific graph structure, a minimum spanning tee. This tree is created using Prim's Algorithm \cite[p.~276]{10.1007/978-3-540-85033-5_27} and the clusters are obtained by cutting edges. With $k$ edges cut the final result is derived through majority voting, a prominent method of this category is Divisive Clustering Ensemble with Automatic Cluster Number (DICLENS) \cite{6035671}. DICLENS is also able to estimate an optimal $k$ by computing all possible results and evaluating their quality.

Locally Adaptive Clustering Algorithms look at different subspaces of the data and compute weights from clustering results in these subspaces \cite{Domeniconi2007}. To create a final result, the computed clusterings and their weights are modeled as a graph, which is then partitioned.

@@

\subsection{Median Partitioning}
The main difference of Median Partitioning in comparison to Object Co-occurrence is that methods in this group tackle consensus clustering as an optimization problem. With the distance between clusterings defined the task of finding a median partition is to look for a solution that would be in the center of the solution space spanned by the base clusterings. In other words, the resulting clustering should have a minimal sum of distances to all base clusterings. As this problem has been shown to be NP-hard \cite{np_median_partition}, heuristics must be used. The different groups of median partitioning methods can be roughly described through the methods that they are based on:
%https://languagetool.org/

@@

Genetic Algorithm based methods require a distance measure and a fitness function to compute a consensus result \cite{Cristofor2002FindingMP}. Through evolutionary steps, different results are combined or modified until the fitness of the result is satisfactory or does not further improve over many iterations. At first, the results to be combined may be picked from the base clusterings or just be random assignments. For the choice of distance measure, Mirkin Distance is often used and comes with additional ideas on how genetic modifications should be applied \cite{5766165}, such that a good result is computed with higher probability. The Sum of Pairwise Distances can also be used with genetic algorithms. The assumption of this measure is that a good consensus result can be found by more strongly considering clusterings far apart in the solution space \cite{6694095}. The idea of using Kernel Functions was also proposed \cite{Vega-Pons:2010:WPC:1786814.1787121}, enabling faster computation regardless of the heuristic used, though mostly genetic algorithms are used with this as well.

Non-negative Matrix Factorization based methods translate the median partitioning problem to a matrix form which can then be solved by running two multiplicative update procedures \cite{Li:2007:SCS:1441428.1442121}. Running those procedures multiple times iteratively improves the result in each iteration and the final result can be obtained by stopping the algorithm at any time or whenever some convergence criterion is met.

Binary Linear Programming based methods create a compact representation of the data and model objects on an intermediate level between points and clusters \cite{HUANG2016131}. For those intermediate objects, binary decisions on cluster ownership are calculated iteratively using Expectation Maximization (EM), whereby within the expectation step of EM the factor graph method \cite{HUANG2016131} is used for estimation. The solution is then mapped back from intermediate objects to data points, assigning the points their labels for the consensus solution.

\part{Theory}
\chapter{Methods}\label{cha:methods}
The newly created tool uses a large number of existing methods that are described in this chapter. Here, I aim to give an overview of what exact methods were used, the theory behind them and in which scenarios they can be/are applied. For a more general overview of some of the topics relating the methods used, refer to the related work of this paper in Chapter \ref{cha:related_work}.

\section{Dimensionality Reduction}\label{sec:dim_reduction}
Dimensionality reduction techniques are able to reduce the number of dimensions in a data-set while trying to maintain as much information as possible. Reducing the number of dimensions is often needed for reducing the computational complexity of clustering and improving the accuracy of the result, but also to visualize the data in a lower-dimensional space. This visualization is then often easier to understand for a user than if all dimensions are shown at once.

\subsection{Principal Component Analysis}
 Principal Component Analysis (PCA) \cite{pca} aims to transform correlated variables into uncorrelated ones thereby omitting less relevant dimensions. Generally, Principal Components (PCs) are linear combinations of the original variables and can be computed through the eigenvectors $V$ of the covariance matrix $C$. To use these PCs for dimensionality reduction, the $k$ ones corresponding to the $k$ largest eigenvalues of the matrix are extracted into a matrix $V_k$ and used as a basis to transform the data. This is done by calculating $data_{reduced}=V_k^T*data_{center}$, whereby the original data must be centered first.

\subsection{t-Distributed Stochastic Neighbor Embedding}
The dimensionality reduction method t-Distributed Stochastic Neighbor Embedding (t-SNE) \cite{Maaten2008VisualizingDU} works by converting the Euclidean distances between points into conditional probabilities that represent similarities. This similarity represents the probability that a point would be picked as the neighbor of another point if neighbors were picked proportionally to their probability density under a Student-t distribution centered at a given point. To layout the points, an initial randomly sampled solution is created and improved on iteratively with gradient descent, such that the similarities are best represented by the points in low-dimensional space. As inputs for the method, the number of dimensions $k$, the perplexity and some optimization parameters are needed. The number of resulting dimensions is defined by $k$ and the optimization parameters can be used to improve the runtime of the algorithm. The perplexity parameter can be seen as a smoothing parameter defining the effective number of neighbors for each point. This parameter is usually also fairly robust \cite{Maaten2008VisualizingDU}, good choices in practice are values between $5$ and $50$.

\subsection{Multi-Dimensional Scaling}
Multi-Dimensional Scaling (MDS) \cite{mds} uses a distance matrix (often called proximity data) and a number of target dimensions $k$ as input. It then tries to lay out the points in a $k$-dimensional space such that the distance between each pair in that space is as close as possible to its true distance. When this method is used with a distance matrix created with Euclidean distances, the result will be equivalent to simply running the PCA method on the data itself.

For the calculation of the solution, MDS builds $n$ $k$-dimensional hypervectors with unknown parameters and aims to minimize the stress value by applying gradient descent. The stress of a configuration is defined as the residual sum of squares for the distances, i.e. the sum of squared errors between the actual distances and the distances in the $k$-dimensional space.

@@

\subsection{k-Means}
K-Means clustering is one of the most prominent methods used for clustering \cite{Jin2010}. The main implementations of k-Means are Lloyd's method \cite{4031353} and MacQueen's method \cite{macqueen1967} although important variants like k-Medians \cite{Dohan2015KmedianAT} and k-Medoids \cite{kmedoids}  are available.

@@

The main difference between Lloyd's and MacQueen's method is that Lloyd's method only recalculates the centroid positions every time an iteration is completed while MacQueen's method also updates them whenever a point changes its assigned clusters. This results in more computational cost during iterations, may allow for faster convergence though. The k-Medians method is different in regards to the distance metric used. Unlike Lloyd's and MacQueen's methods it does not use the Euclidean distance but the Manhattan-distance (1-Norm) to find the centroids. K-Medoids in contrast to all previous methods does not use centroids that may lie anywhere in the data space, but only existing points may be used. These are in this case called medoids and define the objects for which the average dissimilarity to all points within the cluster is minimal.

\subsection{Expectation Maximization}
The Expectation Maximization (EM) algorithm \cite{10.2307/2984875} aims to iteratively find estimates of parameters of a model such that they are assigned the most likely values. To do this, the algorithm alternates between two steps, those being the expectation step and the maximization step. In the expectation step, the expectation of the log-likelihood is defined as a function given the current set of parameters. The maximization step then computes the new set of parameters such that the expected log-likelihood function is maximized.

@@

\subsection{DBSCAN}
Density-Based Spatial Clustering of Application with Noise (DBSCAN) \cite{10.5555/3001460.3001507}, is an algorithm that unlike k-Means is not based on partitioning the data directly but on searching for dense regions in the data. For the calculation, two parameters are needed, namely epsilon ($\epsilon$) and a minimum number of objects ($\mu$ or MinPts). The parameter $\epsilon$ defines how large a region may be considered for neighborhood search and $\mu$ defines how many objects must be within that region for it to be considered dense. The following explanations are needed for a full understanding of the algorithm:
\newline

When defining the neighborhood within the epsilon range for an object $p$ as $N_\epsilon(p)$, the algorithm considers an object $p$ as a core-object if $|N_\epsilon(p)|\geq \mu$.

An object $p$ is said to be directly density-reachable if $p\in N_\epsilon(q) $ and $q$ is a core-object.

@@

Any objects at the border are not core-objects, but will be assigned to a cluster; points outside will be considered noise.
\newline

@@

The output of the OPTICS clustering algorithm is an ordered list of objects with their reachability-distances. To compute this list all points are processed iteratively starting with any point. All points neighboring the current point are updated with their reachability distance from the current point, assigning the minimum of the new and previous values to the point. To select the next point the one with the smallest reachability value is chosen or a new random one if there are none. The resulting list of points with reachability distances can be visualized with a bar chart whereby valleys indicate clusters. As the definition of these valleys can be done in different ways, the result can be considered hierarchical.

\section{Consensus Functions}
Consensus functions have the goal of merging multiple different clustering solutions together such that the overall result represents a consensus between all base clusterings. In this work, I focused on two methods that aim to achieve this.
\subsection{Co-association Matrix based Consensus}
Co-association Matrix based consensus functions \cite{Monti2003} first create a co-association matrix that indicates how often two points $p$ and $q$ appear in a common cluster, i.e. if $p$ and $q$ are clustered together $70\%$ of the time the corresponding entry of the matrix is $0.7$. To extract a consensus result from this matrix different methods can be applied. One possibility is to choose a clustering algorithm that takes this matrix as an input, converts it to a dissimilarity/distance measure and performs clustering with it. Another possibility is to use linking of sets strategies, which is also done for the implementation in the tool. This has the benefit of being trivial to implement and not imposing a cluster structure, as would for example be the case when using k-Means. The possible linking strategies are single-link, using the similarity of the two points of the sets where it is maximal, average-link, using the average similarity between all point pairs, or maximum link, whereby the smallest similarity is used. With each of these strategies the pair of sets for which this linking value is maximal (i.e. they are most similar) is merged. To stop the merging process and obtain the final result either a threshold value can be set for the link value upon which no more merging should be performed or a number of target clusters $k$ can be defined, stopping when only $k$ sets remain. Another option for defining when to stop merging is the Lifetime Criterion. For the implementation of the tool, the average-link strategy is chosen per default with either defining $k$ or using this Lifetime Criterion such that is estimates $k$ itself.

@@

\subsection{Kuhn’s Hungarian method}
Kuhn’s Hungarian method \cite{Kuhn2010} can be used to solve the mathematical assignment problem for the two-dimensional case. A simple example of the assignment problem is to think of $n$ jobs and $n$ workers in a factory that each demand a different payment for each specific job. Now the task is to find the optimal assignment whereby all jobs are performed, all workers are assigned a job and the total cost for all workers is minimal. The assignment problem also allows for reformulation such that the target value is maximized instead of minimized, whenever this better fits the problem and it can be extended to more than two dimensions. In the aforementioned example, an additional dimension could be the tools used, i.e. a worker demands a different payment depending on which task he has to perform with which tool being provided. Also, as an extension to make this more general each dimension may not necessarily need to have the same number of elements. If there are for example more workers than jobs, some workers will not be assigned a job.

@@

\subsection{Kernel Density Estimation}
Kernel Density Estimation aims to estimate the probability density function for a random variable \cite{parzen1962estimation}. Given a sample of points, one is interested in the shape of the unknown density function of the distribution from which the points are drawn. To estimate this function a kernel density estimator is used, requiring a kernel function and a bandwidth parameter. For the kernel function, many different choices are possible, typical ones are Gaussian, multivariate Student and Laplacian kernels \cite{kim2011robust}. The bandwidth parameter defines the smoothness of the resulting function and can be difficult to manually set. For this reason, estimates based on the standard deviation of the samples are available for this parameter \cite{rulethumbdens}.

@@

\part{Implemented Tool}
\chapter{Using the Tool}\label{cha:Tool}
The implemented tool aims to facilitate the computation and selection of good clustering solutions for a given data-set. To do this, both powerful visualizations and algorithms are needed, which are combined here. When working with the created application the user can first import and edit point data, then select clustering algorithms and parameters for meta-clustering and finally use the meta-view to analyze the results or create a consensus results. The final goal after all of this is either having found a clustering solution that is satisfactory or created one using consensus clustering in the final step. This simple workflow can be described with three steps, each with its own view and functionality. In the following sections, the views and possibilities within each view are described. For more details on implementation see Chapter \ref{cha:impl}.

\section{Data Visualization and Manipulation}
To start with the tool, the user needs to import or create a data-set on which the clustering should be performed. As the data may have too many dimensions or include unnecessary columns, it needs to be sanitized as well. For this reason, the first window shown implements functionality for both of these steps. When the application is started a scatter plot without data is shown, as well as a side menu with different buttons. This starting window will further on also be called data-view as it visualizes point data, it can be seen in Figure \ref{fig:data-view}.

@@

The scatter plot in the center-left shows data that was imported or created and allows for changing the shown dimensions by clicking on the center of the corresponding axis. The range of values may also be adjusted this way, clicking on the upper or lower bound of the range on any axis opens a small input box. To automatically set this range, such that the minimum and maximum values of the shown data are just at the border, the small square button in the bottom-left can be clicked. This is also a useful feature for returning to a full view of the data if the range was made smaller to have a zoomed view of the data. The right part of the tool contains the functionality for importing/exporting data, manipulating the data, showing it in a scatter plot matrix and opening the workflow-view which will be explained in the next step.

 Firstly, to import a data-set for visualization and later use the import button in the bottom-right can be selected, opening a file dialog which can load either CSV or ARFF files. These formats are supported as they are most commonly used to store such data. This file dialog can be seen in Figure \ref{fig:data-import}.

@@

After selecting a file the index of cluster labels can be defined if labels should be loaded from the file. Otherwise, the value can be left at $-1$ indicating that no columns should be used for label information. ``$-2$'' may be also be specified, it is used as a shortcut for selecting the last column as cluster label information, removing the need to count columns if it is the last one. The chosen column is then not inserted as a dimension in the data space but used only for the cluster labels, which the color of the loaded points will reflect in the scatter plot of the data-view. In Figure \ref{fig:loaded-data}, one can see how the data-view looks when data with cluster labels is shown. The data-set used for this example is the same as in Section \ref{sec:unob}.

@@

To export data, the export button can be used. This button will open a similar dialog as for importing, though the index of the labels does not need to be defined, as data with cluster labels will be stored such that the labels for all points will be included in the first column. If the data does not contain label information, this first column is completely omitted. The supported format for exporting data is CSV.

Alternatively to importing, the user can also create a data-set himself from within the tool. It is possible to draw points onto the canvas when the ``SinglePoint'' option is selected in the top-right, as in Figure \ref{fig:data-view}. For this, the user only needs to click on the position of the canvas where he wants points placed. It is important to note that if there are more than two dimensions all values for the not shown dimensions will not be set, which may cause a problem for the clustering algorithms later on. Another possibility is the use of the ``ELKIGenerator'', which accepts an XML-style input for defining how to generate data. The definition for this format can be found in corresponding definitions file on the ELKI Github repository \cite{elkixml} or in the tools lib folder. This generator allows defining distributions on dimensions, the number of points for each generated distribution and many additional options like plane rotations. This can be quite useful when wanting to work with a quick sample data-set.
